---
title: "Exercise Manner Prediction Assignment"
author: "Geoffrey Smith"
date: "July 21, 2015"
output: html_document
---

```{r echo=FALSE}
library(knitr)
opts_chunk$set(message=F)
```

##Executive Summary
This report serves as an investigation and analysis of quantified self movement data in an effort to predict the manner in which an exercise has been performed. I will show that using a random forests model we will be able to provide an accurate prediction of the classe.

##Libraries
```{r}
library(caret)
library(corrplot)
library(kernlab)
library(randomForest)
```

##Data processing
Load and clean training data.

```{r}
training <- read.csv("./pml-training.csv", na.strings=c("NA",""," "))
trainingNAs <- apply(training, 2, function(x) {sum(is.na(x))}) #find NAs
trainingClean <- training[,which(trainingNAs == 0)] #remove NAs
trainingClean <- trainingClean[8:length(trainingClean)] #remove first eight identifier columns
```

Load and clean testing data.
```{r}
testing <- read.csv("./pml-testing.csv", na.strings=c("NA",""," "))
testingNAs <- apply(testing, 2, function(x) {sum(is.na(x))}) #find NAs
testingClean <- testing[,which(testingNAs == 0)] #remove NAs
testingClean <- testingClean[8:length(testingClean)] #remove first eight identifier columns
testing <- testingClean

```

Split cleaned training data for cross validation.

```{r}
inTrain <- createDataPartition(y=trainingClean$classe, p=0.7, list=F)
training <- trainingClean[inTrain,]
crossval <- trainingClean[-inTrain,]
```

##Data exploration
Plot correlation matrix to view variable relationships and identify highly correlated predictors.
We can see from the figure below that there are few highly correlated predictors - We will thus include all predictors in the model.

```{r, fig.height=10, fig.width=10}
corrMatrix <- cor(training[, -length(training)])
corrplot(corrMatrix, order="FPC", method="circle", type="lower", tl.cex=0.8, tl.col=rgb(0,0,0))
```

##Model fit
A random forest model will be fitted. Random forests handles unbalanced and low correlated data well.

```{r}
modelFit <- randomForest(classe ~ ., data=training)
modelFit
```

We can see that the error rate estimate is quite low, around 0.5%.
We would expect the out of sample error rate to be similar, if not slightly higher than this.

##Cross validation
We now cross validate the model using a confusion matrix in order to determine the accuracy.

```{r}
predictCrossVal <- predict(modelFit, crossval)
confusionMatrix(crossval$classe, predictCrossVal)
```

The accuracy is around 99.5%, giving an out of sample error of around 0.5%.
This is in line with our previous estimation.

##Prediction
Apply model against testing data set to predict the classe.

```{r}
predictTesting <- predict(modelFit, testing)
predictTesting
```

##Conclusion
The random forest model was able to provide a very accurate prediction of classe based on the use of all of the available parameters.